### 2011-10-07 doc/manuscript.tex
The whole procedure must be rerun.

\section*{Abstract}
Homologous recombination and horizontal gene transfer are investigated with
complete genomes of closely related bacterial species of \textit{Streptococcus
pyogenes} (SPY) and \textit{S.\ dysgalactiae} subspecies \textit{equisimilis}
(SDE).  The former is a pervasive human-specific pathogen, and the latter an
opportunistic human pathogen.  We aims at which parts of the genomic
organization and human pathogenic virulence factors of SDE could have
originated in those of SPY, or vice versa.  Our genome-wide survey of gene
transfer indicated that gene transfer via homologous recombination occurred
relatively from SPY to SDE.  Although we did not find enrichment of virulence
genes in the homologous recombining gene transfer, we found evidence that
virulence genes transferred more often via non-recombining fashion without
replacing parts of destination genomes than did non-virulence genes.  A
genome-wide resource of recombining and non-recombining gene transfer for the
study of the genomes is provided so that research community in pathogenic
bacteria could take advantage of it.

@@

Homologous recombining gene transfer could be studied more readily using the
core genomes of individual bacteria while horizontal gene transfer might be more
easily detected using the pan genomes. Recently, a method of inferring
homologous recombination with whole bacterial genomes was developed
\citep{Didelot2010}. The model of homologous recombination allowed them to study
flows of genetic segment between bacteria individuals of \textit{Bacillus
cereus} species in the core genomes.  Other evolutionary processes acting on
bacterial genomes include genome rearragement, translocation, inversion,
duplication, gain, loss, etc.  These processes did not readily lend themselves
to statistically more sound model-based approaches such as maximum likelihood.
Instead, a parsimony-based approach to inferring gene gain, loss, and transfer
events was developed by \citep{Doyon2011}.  We noted that the first approach
could be used to infer transfer of genomic segments due to homologous
recombination, and the second could do similarly attributed to horizontal gene
transfer.

In short, the complete genomes of the two spcies of SPY and SDE presented us a
unique opportunity to study gene transfer over relatively recent time scales.
We took advantage of the model-based approach for inferring homologous
recombination, and the parsimony-based one for detecting gene duplication, loss,
and horizontal transfer. We compared both of the results in recombining gene
transfer events to show that the parimony-based method inferred recombining gene
transfer events in biological functional categories similar to those inferred
with the model-based approach. This was the first attempt as far as we know to
analyaze SPY and SDE in genome-wide scale for detecting gene transfer events not
only in the core genome but also in the pan genome, which makes the analysis
complete in terms of the amount of used genome data. Our analysis of core
genomes also showed for the first time that the net directionality of
recombining gene transfer from SPY to SDE was stronger than the reverse
direction.

@@

Five genomes except \textit{S.\ equi} ssp.\textit{\ equi} strain 4047 were
chosen (Table S\ref{tab:genome}) to study recombining gene transfer using the
model-based approach.  The genome alignment of the five genomes showed two
distinct genome structures of SPY and SDE (Figure S\ref{fig:mauve}). The SDE and
SDD genomes were larger in length than those of SPY. Although it was unclear
whether SDD was different from either of SDE1 and SDE2 from the graphical view
of the genome alignment, the spcies tree inferred from the genome alignment
showed that SDD branch was attached to the clade branch of SDE1 and SDE2 (Figure
S\ref{fig:tree5}). The distribution of the alignment block lengths was skewed to
the right with mean of 4188 base pairs long (Figure S\ref{fig:blocksize}), which
supported that the model assumption of geometric block length distribution might
be plausible in \texttt{ClonalOrigin} method (see section \textbf{Materials and
Methods} and \citealt{Didelot2010}).  The relative strength ($r/m$) of
recombination compared to mutations indicated that recombination-driven
substitutions affected more sites than did mutation-driven substitutions (Table
S\ref{tab:clonalframe}). This could imply that in the \textit{Streptococcus}
species bacterial recombination process appeared to accelerate molecular
evolution of more sites than did point mutations process.

@@

Along the genome of SDE1 population mutation rate did not appear to change much
(Figure S\ref{fig:scatter3}).  We found a few spots of genomic regions where the
mutation rates were larger than twice the global estimate of mutation rate.  The
recombinant tract length changed in a way that we could not note any particular
positions with notable tract length values compared with other positions along
the genome.  As recombinant tract lengths were smaller, recombination rates per
site seemed larger (Figure S\ref{fig:rhologdelta}).  The recombination rate
appeared to change more dynamically. Especially, a genomic region of SDE1 near
20 kilobase pairs from the origin of replication contained alignment blocks with
bell-shaped recombination rates.

@@

There are 105 possible rooted topologies relating five taxa.  We expected the
dominant topology across sampled genealogies to be concordant with that of the
species tree.  Indeed, 66.7\% of the local gene tree topologies (see section
\textbf{Materials and Methods}) along the alignment blocks were the same as the
species tree, which allowed us to be more confident of the species tree inferred
using the alignment blocks (Table S\ref{tab:Gene-tree-topologies}).  Other
frequently sampled topologies were expected to be potentially informative about
common recombination events.  The second most frequent topology (9.3\%) was one
where SDD was the outgroup, and SPY and SDE were sister clades. Therefore, SPY
and SDE appeared be more closely related with each other than each to SDD's
genome in about tenth of the core genome. The other topologies had less than 4\%
of proportions.

@@

The heat map of Figure \ref{fig:Heatmap-of-recombination} (see Tables
S\ref{tab:heatmap} and S\ref{tab:obsheatmap} for the heat map values and the
observed number of recombinant edges) shows the log base 2 ratios of the
\textit{a posteriori} number of recombinant edges for a pair of donor and
recipient species tree branches and their corresponding expected values given
the species tree (Figure S\ref{fig:tree5}) and the three population parameters
(Table S\ref{tab:three}).  Because recombinant edges from any species tree
branch to the root branch are impossible (see Figure S\ref{fig:clonalorigin}),
the column for the root as recipient was colored green, indicating invalid
recombinant edges.  Similarly, all of the cells for which a recipient branch was
older than the donor, and two branches did not share any contemporary time were
colored green. The diagonal cells were colored gray, indicating no significant
number of recombinant edges, because the observed and expected numbers of
recombinant edges were similar. As recombinant edges starting from and ending in
the same species tree branch would not change the species tree, which would not
change the likelihood much, the cells would be colored gray.  Because SDD is a
strictly veterinary pathogenic bacteria, we did not expect that SDD received
much gene transfer from either SPY1 or SPY2: the log base 2 ratio from SPY1 to
SDD was -2.20, and that from SPY2 to SDD was -1.97.  On the contrary, the
recombing gene transfer from SDE1 to SDD was 0.756, and that from SDE2 to SDD
was 0.698. This relatively higher recombining gene transfer from two SDE
individuals to SDD compared with that from two SPY individuals to SDD reflected
that the SDD was more closely related to SDE than to SPY in most of the genomes,
and the genomic organization between SDD and SDE1/SDE2 were much similar
(Figure S\ref{fig:mauve}).  However, the gene transfer from SDD to either SDE1
or SDE2 was less frequent than the case of reverse direction; SDD effectively
served as a recipient rather than a donor. It was not unclear why SDD received
more genes from SDE1 or SDE2 than did SDE1 or SDE2 from SDD.

Because it was expected that the recombining gene transfer within a clade would
be more frequent than that between clades, not surprisingly the recombining gene
transfer log base 2 ratios between SDE1 and SDE2 (1.17 and 1.20) were about
twice as large as that from SDE1 and SDE2 to SPY1 and SPY2 (0.67, 0.66, 0.60,
and 0.39).  The intensities of recombining gene transfer from external branches
of SDE1 and SDE2 to external branches of SPY1 and SPY2 was similar to those from
SDE1 or SDE2 to SDD (0.76 and 0.70).  After all, either SPY1 or SPY2, and SDD as
recipient appeared to be no different from the view point of SDE1 or SDE2 as
donor since the similar values of recombining gene transfer intensity from SDE1
or SDE2 to the other three external branches.  Surprisingly, the recombining
gene transfer log base 2 ratios between SPY1 and SPY2 (1.06 and 1.07) was
comparable with, but even less than those from SPY1 and SPY2 to SDE1 and SDE2
(1.25, 1.25, 1.17, and 1.29).  Effectively, the recombining gene transfer from a
clade SPY1 or SPY2 to a different clade SDE1 or SDE2 seemed to be no different
from that within the clade of SPY1 and SPY2.  The two external branches of SDE
received genes from either SPY1 or SPY2 via recombining gene transfer as if SDE1
and SDE2 were in the same clade as SPY1 and SPY2.  This was more surprising
because the bacterial recombination between SPY and SDE seemed to happen more
often than that between SDE and SDD even though the genome configurations of SDE
and SDD were more similar than each to SPY.  However, this result could be
concordant with the fact that SPY and SDE shared the same ecological niche
despite their different genomic organization.  The gene transfer log base 2
ratio from SPY1 and SPY2 to SDE branch was also larger than any type of
recombining gene transfer: 1.88 and 1.94, respectively.  Additionally we found
58 genes for which a recombinant edge of certain types were sampled with
probability at least 0.9 by focusing on specific types of recombinant edges
(Table S\ref{tab:genes-transfer}).  Many of the genes in the list were
transferred from SPY to SDE rather than the reverse direction overall.  All
things considered, this epitomized that the recombining gene transfer was biased
toward the net direction from SPY to SDE.

We tried in vain to quantify differential horizontal gene transfer between SPY
and SDE with the parsimony-based approach (Figure \ref{fig:hgt-heatmap}).  Among
the clade-crossing transfers, we did not see any significant enrichment for a
specific direction, as opposed to the result for recombining transfers from the
model-based analysis (Figure S\ref{fig:mowgli-recomb-heatmap}).  SDE1 and SDE2
shared the most events (26 and 34) compared to other sample pairs.  SDD appeared
to acquire many genes from the ancestral SDE branch.  Transfer events between
the SPY and SDE clades were inferred in smaller numbers (<6) compared to those
within each clade (23--34). We were puzzled by this finding because virulence
genes were found to be associated with horizontal gene transfer, and we could
not show net directionality of horizontal gene transfer between SPY and SDE.

@@

Both of the model-based and parimony-based methods used discrepancy between a
reference species tree and gene trees to detect recombining gene transfer. In
essence the two methods worked in relatively similar manner. Firstly, functional
categories that were associated with recombining gene transfer were similar in
both methods (Tables S\ref{tab:functional-all} and
S\label{tab:go-events-recombining}).  Secondly, both of them did not show
association of virulence genes with recombining gene transfer. Lastly, we found
a positive correlation of 0.39 with P-value of $2.2\times10^{-16}$ between the
recombination intensity and the number of recombining gene transfer for the
genes (Figure S\ref{fig:cmpcomowgli}).  The correlation (0.03 with P-value of
0.18) disappeared in the comparison between recombination intensity and number
of horizontal gene transfer events (Figure S\ref{fig:cmpcomowglihgt}).

@@

We created a web-resource for gene transfer available at
\url{http://strep-genome.bscb.cornell.edu}.  For each of the five individual
genomes as a reference genome several annotation tracks were displayed:
alignment blocks, recombination rate per site for each block, local gene tree
topologies, recombination intensity, locations of virulence genes, posterior
probability of recombining gene transfer.  
% To Matt: what could we show in the pan-genome? or from Mowgli's analysis?

@@

Two classes of functional categories notably were associated with recombining
gene transfer.  Firstly, genes involved in sensing queues of environmental
changes would be important for bacteria to adapt to changing environment.  Genes
encoding membrane-bound proteins would be involved in the function.  Several of
functional categories that were associated with gene transfer might be related
with membrane-bound protein coding genes (Table \ref{tab:functional}).  Sugar
phosphotransferase classes were ranked top in the list: phosphrylation must be
important in signal transduction to recognize environmental changes.  The
transport functional categories of carbohydrate transport and sodium ion
transport also belonged to genes encoding membrane-bound proteins.  Also highly
ranked were fatty acid or lipid biosyntheses that must be involved in producing
membrane-bound molecules. These functional categories related with
membrane-bound proteins might allow for bacteria to adapt to changing
environment via recombining gene transfer.

Secondly, aminoacyl tRNA synthetases had been a focus of gene transfer
\citep{Woese2000}, and these aminoacyl tRNA synthetases belonged to the
functional category of translation (GO:0006412).  They were inferred to be 
associated with recombining gene transfer not horizontal gene transfer because
translation category was siginificant only in recombining gene transfer (see
Tables \ref{tab:functional}, \label{tab:go-events} and S\label{tab:go-events-recombining}).

In addition to the two kinds of functional categories, functional categories
concerning ribosomes were also associated with higher recombining gene transfer.
Because genes related with ribosomes were considered to be housekeeping genes,
gene transfer in ribosome genes would be rare, it was unexpected that ribosome
related genes were found to be associated with recombining gene transfer. This
was more significant when we considered all of the recombinant edges (Table
S\ref{tab:functional-all}) not just topology-changing recombinant edges (Table
\ref{tab:functional}) in computing recombination intensity.  Therefore, high
recombination intensity in ribosome related genes would likely to be attributed
to topology-not-changing recombinant edges (Table
S\ref{tab:functional-notopology}).  The parsimony-based approach also showed
enrichment of the recombining transfers in many of the same categories such as
translation, ribosome, and rRNA binding when using all types of recombinant
edges in computing recombination intensity (see Tables
S\ref{tab:go-events-recombining} and S\ref{tab:functional-all}).  Because the
parsimony-based method did not take species tree branch lengths into account, we
were likely to the agreement between the two different methods.

@@

\begin{table}
\caption{\label{tab:functional}Functional categories asscoiated with 
recombination intensity based on topology changing recombinant edge types.}
\begin{tabular}{cccc}
\hline 
P-Value & Count & GO term & Description\tabularnewline
\hline 
1.731e-06 & 38 & GO:0009401 & phosphoenolpyruvate-dependent sugar phosphotransferase system\tabularnewline
2.775e-06 & 33 & GO:0008982 & protein-N(PI)-phosphohistidine-sugar phosphotransferase activity\tabularnewline
4.581e-05 & 22 & GO:0006633 & fatty acid biosynthetic process\tabularnewline
0.0001248 & 20 & GO:0008610 & lipid biosynthetic process\tabularnewline
0.0002675 & 45 & GO:0008643 & carbohydrate transport\tabularnewline
0.000513 & 81 & GO:0006412 & translation\tabularnewline
0.0005617 & 13 & GO:0006814 & sodium ion transport\tabularnewline
0.001019 & 40 & GO:0030529 & ribonucleoprotein complex\tabularnewline
0.001203 & 12 & GO:0031402 & sodium ion binding\tabularnewline
0.00206 & 39 & GO:0003735 & structural constituent of ribosome\tabularnewline
0.002115 & 237 & GO:0008152 & metabolic process\tabularnewline
0.00215 & 26 & GO:0019843 & rRNA binding\tabularnewline
0.00287 & 371 & GO:0016740 & transferase activity\tabularnewline
0.003552 & 66 & GO:0005840 & ribosome\tabularnewline
\hline 
\end{tabular}
\end{table}
\clearpage{}

%
\begin{table}
\caption{\label{tab:go-events}Functional categories asscoiated with 
non-recombining horizontal gene transfer.}
\begin{tabular}{cccc}
\hline 
P-Value & Count & GO term & Description \\
\hline 
7e-61 &  67 & GO:0006313 & transposition, DNA-mediated\\
3e-58 &  59 & GO:0004803 & transposase activity\\
1e-30 &  52 & GO:0015074 & DNA integration\\
1e-28 &  23 & GO:0032196 & transposition\\
1e-13 & 527 & GO:0003677 & DNA binding\\
2e-12 &  86 & GO:0006310 & DNA recombination\\
1e-11 & 197 & GO:0003676 & nucleic acid binding\\
3e-09 &  36 & GO:0016987 & sigma factor activity\\
3e-09 &  36 & GO:0006352 & transcription initiation\\
3e-08 & 142 & GO:0043565 & sequence-specific DNA binding\\
\hline 
\end{tabular}
\end{table}
\clearpage{}

@@

%=============================================================================
% Figure 1. Heat map of recombining gene transfer from ClonalOrigin
\begin{figure}
\includegraphics[scale=0.45]{figures/heatmap-recedge}
\caption{\label{fig:Heatmap-of-recombination}Heatmap of recombination events
using the five genomes.  Each cell of the heat map represents the logarithm base
2 of the number of recombination events inferred relative to its expectation
under the prior, for a donor/recipient pair of branches. The donor is given by
the y-axis; the recipient on the x-axis.  The green cells are meaningless
because the recipient species tree branch of the cells are older than the donor
species branch.}
\end{figure}
\clearpage{}
%=============================================================================

%=============================================================================
% FIGURE 2. - HGT Transfer heatmap
\begin{figure}
\begin{center}
% \includegraphics[width=5in]{figures/trans-hgt-heatmap}
\includegraphics[scale=0.45]{figures/heatmap-hgtedge}
\end{center}
\caption{\label{fig:hgt-heatmap}Distribution of horizontal gene transfers.
Each cell of the color matrix shows the number of inferred non-recombining
horizontal gene transfer events between the different branches of the species
tree.}
\end{figure}
\clearpage{}%
%=============================================================================

@@

@@

%=============================================================================
% Table S. ClonalOrigin's 1st stage result.
\begin{table}
\caption{\label{tab:three}Three population parameter estimates from
ClonalOrigin. 
The estimates of the three parameters were based on one of two independent
MCMC chains.  We considered the similar 
results of the two independent chains as confirmation of 
the convergence of the first stage of \texttt{ClonalOrigin} MCMC.
}
\noindent \begin{centering}
\begin{tabular}{ccc}
& Replicate 1 & Replicate 2\tabularnewline
Parameter & Estimate (IQR$^a$) & Estimate (IQR)\tabularnewline
\hline
$\theta^b$ & 0.081 (0.067, 0.094) & 0.081 (0.067, 0.099)\tabularnewline
$\rho^c$ & 0.012 (0.006, 0.019) & 0.012 (0.006, 0.020)\tabularnewline
$\delta^d$ & 744 (346, 2848) & 723 (348, 2870)\tabularnewline
\hline
\end{tabular}
\par\end{centering}

>>>>

\section*{Abstract}
Homologous recombination and horizontal gene transfer are investigated with
complete genomes of closely related bacterial species of \textit{Streptococcus
pyogenes} (SPY) and \textit{S.\ dysgalactiae} subspecies \textit{equisimilis}
(SDE).  The former is a pervasive human-specific pathogen, and the latter an
opportunistic human pathogen.  We aims at which parts of the genomic
organization and human pathogenic virulence factors of SDE could have originated
in those of SPY, or vice versa.  Our genome-wide survey of gene transfer
indicated that gene transfer via homologous recombination occurred more for the
net direction from SPY to SDE.  Although we did not find enrichment of virulence
genes in homologous recombining gene transfer, we found evidence that
virulence genes transferred more often via non-recombining fashion without
replacing parts of destination genomes than did non-virulence genes.  A
genome-wide resource of recombining and non-recombining gene transfer for the
study of the SPY and SDE genomes is provided so that research community in
pathogenic bacteria could take advantage of it.

@@

Homologous recombining gene transfer could be studied more readily using the
core genomes of individual bacteria while horizontal gene transfer might be more
easily detected using the pan genomes. Recently, a method of inferring
homologous recombination with whole bacterial genomes was developed
\citep{Didelot2010}. The model of homologous recombination allowed them to study
flows of genetic segments between bacteria individuals of \textit{Bacillus
cereus} species in the core genomes.  Other evolutionary processes acting on
bacterial genomes include genome rearragement, translocation, inversion,
duplication, gain, loss, etc.  These processes did not readily lend themselves
to statistically more sound model-based approaches such as maximum likelihood.
Instead, a parsimony-based approach to inferring gene duplication, loss, and
transfer events was developed by \citet{Doyon2011}.  We noted that the first
approach could be used to infer transfer of genomic segments due to homologous
recombination, and the second could help similarly attributed to horizontal gene
transfer.

In short, the complete genomes of the two spcies of SPY and SDE presented us a
unique opportunity to study gene transfer over relatively recent time scales.
We took advantage of the model-based approach for inferring homologous
recombination, and the parsimony-based one for detecting gene duplication, loss,
and horizontal transfer. We compared both of the results in recombining gene
transfer events to show that the parimony-based method inferred recombining gene
transfer events in biological functional categories similar to those inferred
with the model-based approach. This was the first attempt as far as we know to
analyze SPY and SDE in genome-wide scale for detecting gene transfer events not
only in the core genome but also in the pan genome, which makes the analysis
complete in terms of the amount of used genome data. Our analysis of core
genomes also showed for the first time that the net directionality of
recombining gene transfer from SPY to SDE was stronger than the reverse
direction.

@@

Five genomes except \textit{S.\ equi} ssp.\textit{\ equi} strain 4047 were
chosen (Table S\ref{tab:genome}) to study recombining gene transfer using the
model-based approach.  A genome alignment of the five genomes showed two
distinct genome structures of SPY and SDE (Figure S\ref{fig:mauve}). The SDE and
SDD genomes were larger in length than those of SPY. Although it was unclear
whether SDD was different from either of SDE1 and SDE2 from the graphical view
of the genome alignment, a spcies tree inferred from the genome alignment
showed that SDD branch was attached to the clade branch of SDE1 and SDE2 (Figure
S\ref{fig:tree5}). The distribution of the alignment block lengths was skewed to
the right with mean of 4188 base pairs long (Figure S\ref{fig:blocksize}), which
supported that the model assumption of geometric block length distribution might
be plausible in \texttt{ClonalOrigin} method (see section \textbf{Materials and
Methods} and \citealt{Didelot2010}).  The relative strength ($r/m$) of
recombination compared to mutations indicated that recombination-driven
substitutions affected more sites than did mutation-driven substitutions (Table
S\ref{tab:clonalframe}). This could imply that in the \textit{Streptococcus}
species bacterial recombination process appeared to accelerate molecular
evolution of more sites than did point mutations process.

@@

Along the genome of SDE1 population mutation rate did not appear to change much
(Figure S\ref{fig:scatter3}).  We found a few spots of genomic regions where the
mutation rates were larger than twice the global estimate of mutation rate.  The
recombinant tract length changed in a way that we could not note any particular
positions with notable tract length values compared with other positions along
the genome.  As recombinant tract lengths were smaller, recombination rates per
site seemed larger (Figure S\ref{fig:rhologdelta}).  The recombination rate
appeared to change more dynamically. Especially, a genomic region of SDE1 near
200 kilobase pairs from the origin of replication contained alignment blocks with
bell-shaped recombination rates.

@@

There are 105 possible rooted topologies relating five taxa.  We expected the
dominant topology across sampled genealogies to be concordant with that of the
species tree.  Indeed, two thirds of the local gene tree topologies (see section
\textbf{Materials and Methods}) along the alignment blocks were the same as the
species tree, which allowed us to be more confident of the species tree inferred
using the alignment blocks (Table S\ref{tab:Gene-tree-topologies}).  Other
frequently sampled topologies were expected to be potentially informative about
common recombination events.  The second most frequent topology (9.3\%) was one
where SDD was the outgroup, and SPY and SDE were sister clades. Therefore, SPY
and SDE appeared be more closely related with each other than each to SDD's
genome in about tenth of the core genome. The other topologies had less than 4\%
of proportions.

@@

The heat map of Figure \ref{fig:Heatmap-of-recombination} (see Tables
S\ref{tab:heatmap} and S\ref{tab:obsheatmap} for the heat map values and the
observed number of recombinant edges) shows the log base 2 ratios of the
\textit{a posteriori} number of recombinant edges for a pair of donor and
recipient species tree branches and their corresponding expected values given
the species tree (Figure S\ref{fig:tree5}) and the three population parameters
(Table S\ref{tab:three}).  Because recombinant edges from any species tree
branch to the root branch are impossible (see Figure S\ref{fig:clonalorigin}),
the column for the root as recipient was colored green, indicating invalid
recombinant edges.  Similarly, all of the cells for which a recipient branch was
older than the donor, and the donor and recipient branches did not share any contemporary time were
colored green. The diagonal cells were colored gray, indicating no significant
number of recombinant edges, because the observed and expected numbers of
recombinant edges were similar. As recombinant edges starting from and ending in
the same species tree branch would not change the species tree topology, which
might not
change the likelihood much, the cells would be colored gray.  Because SDD is a
strictly veterinary pathogenic bacteria, we did not expect that SDD received
much gene transfer from either SPY1 or SPY2: the log base 2 ratio from SPY1 to
SDD was -2.20, and that from SPY2 to SDD was -1.97.  On the contrary, the
recombing gene transfer from SDE1 to SDD was 0.756, and that from SDE2 to SDD
was 0.698. This relatively higher recombining gene transfer from two SDE
individuals to SDD compared with that from two SPY individuals to SDD reflected
that the SDD was more closely related to SDE than to SPY in most of the genomes,
and the genomic organization between SDD and SDE1/SDE2 were much similar
(Figure S\ref{fig:mauve}).  However, the gene transfer from SDD to either SDE1
or SDE2 was less frequent than the case of reverse direction; SDD effectively
served as a recipient rather than a donor. It was not unclear why SDD received
more genes from SDE1 or SDE2 than did SDE1 or SDE2 from SDD.

Because it was expected that the recombining gene transfer within a clade would
be more frequent than that between clades, not surprisingly the recombining gene
transfer log base 2 ratios between SDE1 and SDE2 (1.17 and 1.20) were about
twice as large as that from SDE1 and SDE2 to SPY1 and SPY2 (0.67, 0.66, 0.60,
and 0.39).  The intensities of recombining gene transfer from external branches
of SDE1 and SDE2 to external branches of SPY1 and SPY2 was similar to those from
SDE1 or SDE2 to SDD (0.76 and 0.70).  After all, either SPY1 or SPY2, and SDD as
recipient appeared to be no different from the view point of SDE1 or SDE2 as
donor since the similar values of recombining gene transfer intensity from SDE1
or SDE2 to one of the other three external branches. Surprisingly, the recombining
gene transfer log base 2 ratios between SPY1 and SPY2 (1.06 and 1.07) was
comparable with, yet even less than those from SPY1 and SPY2 to SDE1 and SDE2
(1.25, 1.25, 1.17, and 1.29).  Effectively, the recombining gene transfer from a
clade SPY1 or SPY2 to a different clade SDE1 or SDE2 seemed to be no different
from that within the clade of SPY1 and SPY2.  The two external branches of SDE
received genes from either SPY1 or SPY2 via recombining gene transfer as if SDE1
and SDE2 were in the same clade as SPY1 and SPY2.  This was more surprising
because the bacterial recombination between SPY and SDE seemed to happen more
often than that between SDE and SDD even though the genome configurations of SDE
and SDD were more similar than each to SPY.  However, this result could be
concordant with the fact that SPY and SDE shared the same ecological niche
despite their dissimilar genomic organization.  The gene transfer log base 2
ratio from SPY1 and SPY2 to SDE branch was also larger than any type of
recombining gene transfer: 1.88 and 1.94, respectively.  Additionally we found
58 genes for which a recombinant edge of certain types were sampled with
probability at least 0.9 by focusing on specific types of recombinant edges
(Table S\ref{tab:genes-transfer}).  Many of the genes in the list were
transferred from SPY to SDE rather than the reverse direction overall.  All
things considered, this epitomized that the recombining gene transfer was biased
toward the net direction from SPY to SDE.

We tried in vain to quantify differential horizontal gene transfer between SPY
and SDE with the parsimony-based approach (Figure \ref{fig:hgt-heatmap}).  Among
the clade-crossing transfers, we did not see any significant enrichment for a
specific direction, as opposed to the result for recombining transfers from the
model-based analysis (Figure S\ref{fig:mowgli-recomb-heatmap}).  SDE1 and SDE2
shared the most events (26 and 34) compared to other sample pairs.  SDD appeared
to acquire many genes from the ancestral SDE branch.  Transfer events between
the SPY and SDE clades were inferred in smaller numbers (<6) compared to those
within each clade. We were puzzled by this finding because virulence
genes were found to be associated with horizontal gene transfer, and yet we could
not show net directionality of horizontal gene transfer between SPY and SDE.

@@

Both of the model-based and parimony-based methods used discrepancy between a
reference species tree and gene trees to detect recombining gene transfer. In
essence the two methods worked in relatively similar manner. Firstly, functional
categories that were associated with recombining gene transfer were similar in
both methods (Tables S\ref{tab:functional-all} and
S\ref{tab:go-events-recombining}).  Secondly, both of them did not show
association of virulence genes with recombining gene transfer. Lastly, we found
a positive correlation of 0.39 with P-value of $2.2\times10^{-16}$ between the
recombination intensity and the number of recombining gene transfer for the
genes (Figure S\ref{fig:cmpcomowgli}).  The correlation (0.03 with P-value of
0.18) disappeared in the comparison between recombination intensity and number
of horizontal gene transfer events (Figure S\ref{fig:cmpcomowglihgt}).

@@

We created a web-resource (Figure S\ref{fig:ucsc}) for gene transfer available at
\url{http://strep-genome.bscb.cornell.edu}.  For each of the five individual
genomes as a reference genome several annotation tracks were displayed:
alignment blocks, recombination rate per site for each block, local gene tree
topologies, recombination intensity, locations of virulence genes, posterior
probability of recombining gene transfer.  
% To Matt: what could we show in the pan-genome? or from Mowgli's analysis?

@@

Two classes of functional categories were notably to be inferred to associated
with recombining gene transfer.  Firstly, genes involved in sensing queues of
environmental changes would be important for bacteria to adapt to changing
environment.  Genes encoding membrane-bound proteins would be involved in the
function.  Several of functional categories that were associated with gene
transfer might be related with membrane-bound protein coding genes (Table
\ref{tab:functional}).  Sugar phosphotransferase classes were ranked top in the
list: phosphrylation could be important in signal transduction to recognize
environmental changes.  The transport functional categories of carbohydrate
transport and sodium ion transport also belonged to genes encoding
membrane-bound proteins.  Also highly ranked were fatty acid and lipid
biosyntheses that must be involved in producing membrane-bound protein
molecules. These functional categories related with membrane-bound proteins
might allow for bacteria to adapt to changing environment via recombining gene
transfer.  Secondly, aminoacyl tRNA synthetases had been a focus of gene
transfer \citep{Woese2000}, and these aminoacyl tRNA synthetases belonged to the
functional category of translation (GO:0006412).  They were inferred to be
associated with recombining gene transfer not horizontal gene transfer because
translation category was siginificant only in recombining gene transfer (see
Tables \ref{tab:functional}, \ref{tab:go-events} and
S\ref{tab:go-events-recombining}).

In addition to the two kinds of functional categories, functional categories
concerning ribosomes were also associated with higher recombining gene transfer.
Because genes related with ribosomes were considered to be housekeeping genes,
gene transfer in ribosome genes would be rare, it was unexpected that ribosome
related genes were found to be associated with recombining gene transfer. This
was more significant when we considered all of the recombinant edges (Table
S\ref{tab:functional-all}) not just topology-changing recombinant edges (Table
\ref{tab:functional}) in computing recombination intensity.  Therefore, high
recombination intensity in ribosome related genes would be likely attributed
to topology-not-changing recombinant edges (Table
S\ref{tab:functional-notopology}).  The parsimony-based approach also showed
enrichment of the recombining transfers in many of the same categories such as
translation, ribosome, and rRNA binding when using all types of recombinant
edges in computing recombination intensity (see Tables
S\ref{tab:functional-all} and S\ref{tab:go-events-recombining}).  Because the
parsimony-based method did not take species tree branch lengths into account, 
the agreement between the two different methods might be likely.

@@

\begin{table}
\caption{\label{tab:functional}Functional categories asscoiated with 
recombination intensity based on topology changing recombinant edge types.}
\begin{tabular}{cccl}
\hline 
P-Value & Count & GO term & Description\tabularnewline
\hline 
1.731e-06 & 38 & GO:0009401 & phosphoenolpyruvate-dependent sugar phosphotransferase system\tabularnewline
2.775e-06 & 33 & GO:0008982 & protein-N(PI)-phosphohistidine-sugar phosphotransferase activity\tabularnewline
4.581e-05 & 22 & GO:0006633 & fatty acid biosynthetic process\tabularnewline
0.0001248 & 20 & GO:0008610 & lipid biosynthetic process\tabularnewline
0.0002675 & 45 & GO:0008643 & carbohydrate transport\tabularnewline
0.000513 & 81 & GO:0006412 & translation\tabularnewline
0.0005617 & 13 & GO:0006814 & sodium ion transport\tabularnewline
0.001019 & 40 & GO:0030529 & ribonucleoprotein complex\tabularnewline
0.001203 & 12 & GO:0031402 & sodium ion binding\tabularnewline
0.00206 & 39 & GO:0003735 & structural constituent of ribosome\tabularnewline
0.002115 & 237 & GO:0008152 & metabolic process\tabularnewline
0.00215 & 26 & GO:0019843 & rRNA binding\tabularnewline
0.00287 & 371 & GO:0016740 & transferase activity\tabularnewline
0.003552 & 66 & GO:0005840 & ribosome\tabularnewline
\hline 
\end{tabular}
\end{table}
\clearpage{}

%
\begin{table}
\caption{\label{tab:go-events}Functional categories asscoiated with 
non-recombining horizontal gene transfer.}
\begin{tabular}{cccl}
\hline 
P-Value & Count & GO term & Description \\
\hline 
7e-61 &  67 & GO:0006313 & transposition, DNA-mediated\\
3e-58 &  59 & GO:0004803 & transposase activity\\
1e-30 &  52 & GO:0015074 & DNA integration\\
1e-28 &  23 & GO:0032196 & transposition\\
1e-13 & 527 & GO:0003677 & DNA binding\\
2e-12 &  86 & GO:0006310 & DNA recombination\\
1e-11 & 197 & GO:0003676 & nucleic acid binding\\
3e-09 &  36 & GO:0016987 & sigma factor activity\\
3e-09 &  36 & GO:0006352 & transcription initiation\\
3e-08 & 142 & GO:0043565 & sequence-specific DNA binding\\
\hline 
\end{tabular}
\end{table}
\clearpage{}

@@

%=============================================================================
% Figure 1. Heat map of recombining gene transfer from ClonalOrigin
\begin{figure}
\includegraphics[scale=0.45]{figures/heatmap-recedge}
\caption{\label{fig:Heatmap-of-recombination}Heatmap of recombination events
using the five genomes.  Each cell of the heat map represents the logarithm base
2 of the number of recombination events inferred relative to its expectation
under the prior, for a donor and recipient pair of branches. The donor is given by
the y-axis; the recipient on the x-axis.  The green colored cells are meaningless
because the recipient species tree branch of the cells are older than the donor
species branch.}
\end{figure}
\clearpage{}
%=============================================================================

%=============================================================================
% FIGURE 2. - HGT Transfer heatmap
\begin{figure}
\begin{center}
% \includegraphics[width=5in]{figures/trans-hgt-heatmap}
\includegraphics[scale=0.45]{figures/heatmap-hgtedge}
\end{center}
\caption{\label{fig:hgt-heatmap}Heat map of horizontal gene transfers using the
six genomes.
Each cell of the color matrix shows the number of inferred non-recombining
horizontal gene transfer events between the different branches of the species
tree. Black colored cells represent no inferred events.}
\end{figure}
\clearpage{}%
%=============================================================================

@@

%=============================================================================
% Table S6. ClonalOrigin's 1st stage result.
\begin{table}
\caption{\label{tab:three}Three population parameter estimates from
ClonalOrigin. 
The estimates of the three parameters were based on one of two independent
MCMC chains.  We considered the similar 
results of the two independent chains as confirmation of 
the convergence of the first stage of \texttt{ClonalOrigin} MCMC.
}
\noindent \begin{centering}
\begin{tabular}{ccc}
\hline
& Replicate 1 & Replicate 2\tabularnewline
Parameter & Estimate (IQR$^a$) & Estimate (IQR)\tabularnewline
\hline
$\theta^b$ & 0.081 (0.067, 0.094) & 0.081 (0.067, 0.099)\tabularnewline
$\rho^c$ & 0.012 (0.006, 0.019) & 0.012 (0.006, 0.020)\tabularnewline
$\delta^d$ & 744 (346, 2848) & 723 (348, 2870)\tabularnewline
\hline
\end{tabular}
\par\end{centering}

@@
